Open-source tools in R for landscape ecology

### 2021-02-02 paper/paper.Rmd
Clean references

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- R packages
- reproducibility
- scientific software
abstract: |
  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology. **Recent findings** In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of *R* users within the landscape ecology community. **Summary** We found that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

API & Application programming interface. A set of protocols for sending and retrieving information from a server \\
cell & Smallest, rectangular unit of raster data model \\
CRAN & Comprehensive R Archive Network. A repository for *R* package \\
CRS & Coordinate references systems. A coordinate-based local, regional or global system used to locate geographical entities \\
EPSG & Public registry of geodetic datums, spatial reference systems, Earth ellipsoids, coordinate transformations and related units of measurement \\
GDAL & Open source translator library for raster and vector geospatial data formats \\
GIS & Geographical information system \\
landscape & Mosaic of land covers, ecosystems, habitat types, land uses \\
NLM & Neutral landscape model \\
open-source software & Software released under licenses that allow to use, modify and distribute it \\
patch & Neighboring cells with same value \\
PROJ & Open source generic coordinate transformation software \\
raster & Cell grid based data model \\
resolution & Size in meter or degree of one cell \\
*R* package & User-created software extension to *R* programming language \\
SDM & Species distribution modeling \\
Simple Features & A set of standards that specify a common storage and access model of geographic features \\
vector & Points, lines, and polygons-based data model \\

@@

Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes can vary spatially \cite{With2019}.
The importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance today \cite{Simova2012,Estes2018}.
Thus, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize spatial data.

@@

One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}.
Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2019,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.

@@

While *R* has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
Importantly, main *R* packages for spatial data use the external *GDAL* \cite{GDAL} and *PROJ* \cite{PROJ} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
Currently, the most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
A *raster* successor, *terra*, aimed at the simpler interface and improved performance is being developed \cite{R-terra}, however, it could take several years for this package to be adopted by other developers and users.
Alternatively, the *stars* package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
Additionally, there are packages that improve some basic raster operations in terms of computational performance or compatibility between raster and vector operations, such as *fasterize* \cite{Ross2020}, *rasterDT* \cite{OBrien2020}, or *exactextractr* \cite{Baston2020}.

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The *sp* package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 *R* packages directly depend or imports *sp*.
However, *sp* is not actively developed anymore, and its recommended successor is the *sf* package \cite{Pebesma2018}.
Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019a}.
*sf* builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Nowadays, spatial data at many scales is available from an abundance of online-accessible sources.
A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an *R* session.
Since publicly available data is becoming more prominent, so are *R* packages to access them.
Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database to download region and country data, the *elevatr* package to access raster elevation data \cite{R-elevatr}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{R-sen2r} to download Sentinel-2 optical images.
Also, the `getData()` function from the *raster* package allows users to download climatic and bioclimatic data from WorldClim v1.4.
Additionally, the *rgee* package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.

@@

Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
Also the popular plotting package *ggplot2* \cite{Wickham2016a} has an extension especially designed for plotting spatial data named *ggspatial* \cite{R-ggspatial}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Quick interactive visualization of spatial data can be done with the *mapview* package\cite{R-mapview}.
Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.
A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

@@

The introduction of the *FRAGSTATS* in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the *landscapemetrics* package \cite{Hesselbarth2019a} allows to calculate the most widely used landscape metrics in within the *R* environment.

More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
The *geodiv* package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
Additionally, the *belg* package allows calculation of the Boltzmann entropy of a landscape gradient \cite{R-belgpaper}.

Most landscape metrics are a represented by a single number depicting specific characteristics of a local landscape.
Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
<!-- MH: Do we have a reference here @Jakub? -->
Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g. Euclidian, Manhattan, Mahlanobis).
<!-- LJG added some examples, but not 100% if they're the distances you're thinking of -->
This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
All of the spatial signatures methods mentioned above are implemented in the *motif* package \cite{R-motif}.

@@

Point pattern analysis uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
The *spatstat* package \cite{R-spatstat} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics. 
A comprehensive textbook covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.

Distance-based methods allow to detect and correct for spatial autocorrelation in data.
It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling. 
The *spdep* package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
The *rinla* \cite{rue2009approximate} (\url{https://www.r-inla.org}) and *inlabru* \cite{R-inlabru} packages also provide functionality for modeling of spatially structured data.

Finally, the spatial structure of continuous landscapes can be quantified and compared with geostatistical tools, such as variograms and correlograms. 
R packages *geoR* \cite{R-geoR} and *gstat* \cite{R-gstat} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
Geostatistics also allows for spatial data simulations.
<!-- MH: I am not sure if I understand the last sentence, but is it worth to reference the NLM section here? -->

@@

Species distribution modeling (SDM) examines how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species' distributions, mainly to infer ecological processes and predict future species' distributions \cite{Wiersma2011}.
Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010,Norberg2019}.
Additionally, textbooks introducing basic concepts of SDM in *R* exists \cite{Guisan2017,Fletcher2019}.

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs.
Popular approaches and packages include generalized linear models using, e.g., the *stats* \cite{RCoreTeam2019} package; generalized additive models using, e.g., the *mgcv* \cite{Wood2017} or *lme4* \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the *rpart* \cite{Therneau2019}, *randomForest* \cite{Liaw2002} or *ranger* \cite{Wright2017} package or multivariate data analysis using , e.g., the *ade4* \cite{Dray2007} or *vegan* \cite{Oksanen2019} package.
Also, packages specifically designed for SDM exist, including the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020}, *PresenceAbsence* \cite{Freeman2008}, or *zoon* \cite{Golding2018} packages. 
Additionally, packages such as *ENMeval* \cite{R-enmeval} provide functionality for model tuning and evaluation.

Related to SDMs, there is a growing number of *R* packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
Widely used *R* packages include *ctmm* \cite{calabrese2016} and *adehabitatHR* \cite{Calenge2006} for home-range estimation, *moveHMM* \cite{michelot2016} for the classification of behavioral states, and *amt* \cite{Signer2019} for habitat selection.
A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{joo2020}.

@@

Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}.
Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}.
While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species such as habitat associations and dispersal distances \cite{Tischendorf2000,With2019}.
Given its broad concept, many different measures of connectivity exist \cite{Kindlmann2008}.
At the patch level, structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g., amount of suitable habitat) \cite{Kindlmann2008,With2019}.
Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}).
Furthermore, the *lconnect* package \cite{Mestre2019} and *Makurhini* package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
Another way to describe connectivity is based on graph theory with the advantage that functional connectivity can also be included \cite{Kindlmann2008}.
In graph theory \cite{Laita2011}, landscapes are described by nodes (i.e., habitat patches) connected by and functional connections called links (or edges) \cite{Laita2011}.
The *grainscape* package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
More generally, the *igraph* package \cite{Csardi2006} provides functionality for graph theoretic analyses.
Resistance surfaces and least-cost paths are other tools to model functional connectivity which include attributes of the matrix.
The resistance surface describes the effects of facilitating or hindering the landscape's characteristics for an organism moving within it \cite{Adriaensen2003}.
Least-cost paths can be calculated using the *gdistance* package \cite{vanEtten2017}.
Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019a}, and is provided by the recently published *samc* package \cite{Marx2020}.

@@

& terra & Raster (and some vector) data handling and analysis & \cite{R-terra} \\

@@

& elevatr & Download elevation data & \cite{R-elevatr} \\
& rgbif & Download biodiversity data & \cite{R-rgbif} \\

@@

& FedData & Download geospatial data from US federal sources & \cite{R-FedData} \\

@@

& sen2r & Download satellite data from Sentinel & \cite{R-sen2r} \\

@@

& mapview & Interactive viewing of spatial data & \cite{R-mapview} \\
& leaflet & Create interactive web maps & \cite{R-leaflet} \\

@@

\multirow{4}{*}{\shortstack[l]{Quantifying landscape\\characteristics}} & landscapemetrics & Quantify categorical landscape patterns & \cite{Hesselbarth2019a} \\
& belg & Calculate the Boltzmann entropy & \cite{R-belgpaper} \\
& motif & Pattern-based spatial analysis & \cite{R-motif} \\

@@

>>>>

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- R packages
- reproducibility
- scientific software
abstract: |
  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology. **Recent findings** In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of *R* users within the landscape ecology community. **Summary** In this we find that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually updating the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

API & Application programming interface. A set of protocols for sending and retrieving information from a server \\
cell & Smallest, rectangular unit of raster data model \\
CRAN & Comprehensive R Archive Network. A repository for *R* package \\
CRS & Coordinate references systems. A coordinate-based local, regional or global system used to locate geographical entities \\
EPSG & Public registry of geodetic datums, spatial reference systems, Earth ellipsoids, coordinate transformations and related units of measurement \\
GDAL & Open source translator library for raster and vector geospatial data formats \\
GIS & Geographical information system \\
landscape & Mosaic of land covers, ecosystems, habitat types, land uses \\
NLM & Neutral landscape model \\
open-source software & Software released under licenses that allow to use, modify and distribute it \\
patch & Neighboring cells with same value \\
PROJ & Open source generic coordinate transformation software \\
raster & Cell grid based data model \\
resolution & Size in meter or degree of one cell \\
R package & User-created software extension to R programming language \\
SDM & Species distribution modelling \\
Simple Features & A set of standards that specify a common storage and access model of geographic features \\
vector & Points, lines, and polygons-based data model \\

@@

Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is important and ecological processes can vary spatially \cite{With2019}.
The importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance today \cite{Simova2012,Estes2018}.
Thus, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize spatial data.

@@

One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}.
Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
The expanding CRAN Task Views (curated lists of packages related to a certain topic) document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Wegmann2016,Fletcher2018,Lovelace2019,Pebesma2019a}.
A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.

@@

While *R* has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data.
However, because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many *R* packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the *R* language.
Importantly, main *R* packages for spatial data use the external *GDAL* \cite{GDAL/OGRcontributors2020} and *PROJ* \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, *R* allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
Currently, the most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
A *raster* successor, *terra*, aimed at the simpler interface and improved performance is being developed \cite{Hijmans2021}, however, it could take several years for this package to be adopted by other developers and users.
Alternatively, the *stars* package can be used to read and process raster data focusing on spatial-temporal data cubes \cite{Pebesma2019}.
Additionally, there are packages that improve some basic raster operations in terms of computational performance or compatibility between raster and vector operations, such as *fasterize* \cite{Ross2020}, *rasterDT* \cite{OBrien2020}, or *exactextractr* \cite{Baston2020}.

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The *sp* package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 *R* packages directly depend or imports *sp*.
However, *sp* is not actively developed anymore, and its recommended successor is the *sf* package \cite{Pebesma2018}.
Besides many advantages and strength of the *sf* package in terms of spatial data handling, it also integrates into the widely used *tidyverse* packages \cite{Wickham2019}.
*sf* builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Nowadays, spatial data at many scales is available from an abundance of online-accessible sources.
A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an *R* session.
Since publicly available data is becoming more prominent, so are *R* packages to access them.
Packages include *rnaturalearth* \cite{South2017} to access the Natural Earth database to download region and country data, the *elevatr* package to access raster elevation data \cite{Hollister2020}, the *rgbif* package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the *BIEN* package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the *FedData* package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the *getlandsat* package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the *MODIS* package \cite{Mattiuzzi2020} to download MODIS products, and *sen2r* \cite{Ranghetti2020} to download Sentinel-2 optical images.
Also, the `getData()` function from the *raster* package allows users to download climatic and bioclimatic data from WorldClim v1.4.
Additionally, the *rgee* package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from *R* using dedicated packages, such as *rgrass7* \cite{Bivand2021} for GRASS GIS, *RSAGA* \cite{Brenning2018} for SAGA GIS, and *rgee* \cite{Aybar2020} for Google Earth Engine.

@@

Further plotting methods for raster objects can be found in the *rasterVis* package \cite{Lamigueiro2020}.
Also the popular plotting package *ggplot2* \cite{Wickham2016} has an extension especially designed for plotting spatial data named *ggspatial* \cite{Dunnington2020}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the *cartography* package \cite{Giraud2016}.
The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Quick interactive visualization of spatial data can be done with the *mapview* package\cite{Appelhans2020}.
Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{Cheng2021}.
A slightly different approach to visualizing spatial data is adapted by the *rayshader* package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

@@

The introduction of the *FRAGSTATS* in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the *landscapemetrics* package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics in within the *R* environment.

More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}.
The *geodiv* package \cite{Smith2020} allows calculation of gradient surface metrics to facilitate continuous analysis of landscape features.
Additionally, the *belg* package allows calculation of the Boltzmann entropy of a landscape gradient \cite{Nowosad2020}.

Most landscape metrics are a represented by a single number depicting specific characteristics of a local landscape.
Another possibility is to derive spatial signatures - a multi-value representation of landscape composition and configuration, such as a co-occurrence histogram.
<!-- MH: Do we have a reference here @Jakub? -->
Spatial signatures calculated for many landscapes can be compared using one of a set of existing distance measures (e.g. Euclidian, Manhattan, Mahlanobis).
<!-- LJG added some examples, but not 100% if they're the distances you're thinking of -->
This enables several types of spatial analysis on categorical raster data, such as searching for similar landscapes, detecting changes between landscape patterns, and spatial clustering of landscapes based on their composition and configuration.
All of the spatial signatures methods mentioned above are implemented in the *motif* package \cite{Nowosad2021}.

@@

Point pattern analysis uses event-level data, such as locations of individuals, and links the spatial pattern to the ecological process.
The *spatstat* package \cite{Baddeley2005} contains functionality for point pattern analysis, including exploratory analysis; simulation of point process models; and modeling fitting, inference, and diagnostics. 
A comprehensive textbook covering both theoretical background as well as applied examples can be found here \cite{Baddeley2015}.

Distance-based methods allow to detect and correct for spatial autocorrelation in data.
It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling. 
The *spdep* package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
The *rinla* \cite{Rue2009} (\url{https://www.r-inla.org}) and *inlabru* \cite{Bachl2019} packages also provide functionality for modeling of spatially structured data.

Finally, the spatial structure of continuous landscapes can be quantified and compared with geostatistical tools, such as variograms and correlograms. 
R packages *geoR* \cite{Diggle2007} and *gstat* \cite{Pebesma2004} provide functionality for this type of analysis, as well as interpolation methods, known as (co-)kriging.
Geostatistics also allows for spatial data simulations.
<!-- MH: I am not sure if I understand the last sentence, but is it worth to reference the NLM section here? -->

@@

Species distribution modeling (SDM) examines how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species' distributions, mainly to infer ecological processes and predict future species' distributions \cite{Wiersma2011}.
Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010,Norberg2019}.
Additionally, textbooks introducing basic concepts of SDM in *R* exists \cite{Guisan2017,Fletcher2018}.

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2018}, there is also a large number of *R* packages used for SDMs.
Popular approaches and packages include generalized linear models using, e.g., the *stats* \cite{RCoreTeam2019} package; generalized additive models using, e.g., the *mgcv* \cite{Wood2017} or *lme4* \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the *rpart* \cite{Therneau2019}, *randomForest* \cite{Liaw2002} or *ranger* \cite{Wright2017} package or multivariate data analysis using , e.g., the *ade4* \cite{Dray2007} or *vegan* \cite{Oksanen2019} package.
Also, packages specifically designed for SDM exist, including the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020}, *PresenceAbsence* \cite{Freeman2008}, or *zoon* \cite{Golding2018} packages. 
Additionally, packages such as *ENMeval* \cite{Muscarella2014} provide functionality for model tuning and evaluation.

Related to SDMs, there is a growing number of *R* packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{Signer2015}, behavioral states \cite{Langrock2012}, or habitat selection \cite{Fieberg2020}.
Widely used *R* packages include *ctmm* \cite{Calabrese2016} and *adehabitatHR* \cite{Calenge2006} for home-range estimation, *moveHMM* \cite{Michelot2016} for the classification of behavioral states, and *amt* \cite{Signer2019} for habitat selection.
A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{Joo2020}.

@@

Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}.
Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}.
While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species such as habitat associations and dispersal distances \cite{Tischendorf2000,With2019}.
Given its broad concept, many different measures of connectivity exist \cite{Kindlmann2008}.
At the patch level, structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g., amount of suitable habitat) \cite{Kindlmann2008,With2019}.
Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}).
Furthermore, the *lconnect* package \cite{Mestre2019} and *Makurhini* package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
Another way to describe connectivity is based on graph theory with the advantage that functional connectivity can also be included \cite{Kindlmann2008}.
In graph theory \cite{Laita2011}, landscapes are described by nodes (i.e., habitat patches) connected by and functional connections called links (or edges) \cite{Laita2011}.
The *grainscape* package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
More generally, the *igraph* package \cite{Csardi2006} provides functionality for graph theoretic analyses.
Resistance surfaces and least-cost paths are other tools to model functional connectivity which include attributes of the matrix.
The resistance surface describes the effects of facilitating or hindering the landscape's characteristics for an organism moving within it \cite{Adriaensen2003}.
Least-cost paths can be calculated using the *gdistance* package \cite{vanEtten2017}.
Absorbing Markov chains quantify landscape connectivity as the combination of movement and mortality based on the landscape characteristics \cite{Fletcher2019}, and is provided by the recently published *samc* package \cite{Marx2020}.

@@

& terra & Raster (and some vector) data handling and analysis & \cite{Hijmans2021} \\

@@

& elevatr & Download elevation data & \cite{Hollister2020} \\
& rgbif & Download biodiversity data & \cite{Chamberlain2017} \\

@@

& FedData & Download geospatial data from US federal sources & \cite{Bocinsky2019} \\

@@

& sen2r & Download satellite data from Sentinel & \cite{Ranghetti2020} \\

@@

& mapview & Interactive viewing of spatial data & \cite{Appelhans2020} \\
& leaflet & Create interactive web maps & \cite{Cheng2021} \\

@@

\multirow{4}{*}{\shortstack[l]{Quantifying landscape\\characteristics}} & landscapemetrics & Quantify categorical landscape patterns & \cite{Hesselbarth2019} \\
& belg & Calculate the Boltzmann entropy & \cite{Nowosad2020} \\
& motif & Pattern-based spatial analysis & \cite{Nowosad2021} \\

@@

### Important references
